---
title: "QuantSeq Data Exploratory Analysis"
output:
  html_notebook:
    code_folding: hide
    toc: yes
    toc_float: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(plotly)
```


## Quantseq Dataset and Metadata

Quantseq data set

```{r}
getwd()
dat <- read.csv("../raw-data/coinfection 191 samples QuantSeq matched metadata PHGC16 output from Bluebee ssc10.2.csv")
dim(dat)
head(colnames(dat))
head(rownames(dat))
head(dat$X)
head(levels(dat$X))
length(levels(dat$X))
x <- plyr::laply(strsplit(as.character(dat$X), split = "__"), function(ss) length(ss))

dat <- dat[-which(x==2),]
rownames(dat) <- dat$X
dat <- dat[, -1]
colnames(dat) <- gsub(pattern = "X", "", colnames(dat))
dat <- dat[, -which(log(apply(dat, 2, sum)) <= 10)] 
# dat[which(apply(dat, 1, sum)/sum(dat)>.01),]
dat <- dat[-which(apply(dat, 1, sum)/sum(dat)>.01)[1:3],] # delete hemoglobin genes
# barplot(apply(dat, 1, sum)/sum(dat))
# dat[which(apply(dat, 1, sum)/sum(dat) >.05),]
```




Metadata

```{r}
metadata <- read.csv("../raw-data/coinfectionPHGC16metadata191samples (1).csv")
metadata <- metadata%>%dplyr::arrange(RNAid) %>% 
  dplyr::mutate(RNAid = as.character(RNAid), 
                Plate = as.factor(Plate)) %>% 
  dplyr::filter(RNAid %in%colnames(dat))
dat <- dat[, metadata$RNAid]
# check colnames of dat and metadata are the same
all(colnames(dat)==metadata$RNAid)
alldat <- list(counts = dat, metadata = metadata)
saveRDS(alldat, file = paste0("../data/dat.rds"))
```
How many samples per each combination of WUR and Vac Status?
```{r}
library(dplyr)
str(metadata)
table(metadata$VacStatue, metadata$WUR, metadata$dpv)

num.sample <- metadata %>% 
  dplyr::mutate(WUR_Vac = paste0(WUR, "_", VacStatue), 
                Litters = as.character(gsub("([0-9]+).*$", "\\1", Litter)),
                PHGC_16_PigID = as.factor(PHGC_16_PigID)) %>%
  dplyr::mutate(WUR_Vac = factor(WUR_Vac, levels = c( "AA_nonVaVac", "AB_nonVaVac", "AA_Vac", "AB_Vac"))) #%>%
  #dplyr::filter(RINe >=7)

ggplot2::ggplot(dat =num.sample, aes(x = dpv)) +
  ggplot2::geom_bar(aes(fill = WUR_Vac))+
  ggplot2::scale_x_continuous(breaks=unique(num.sample$dpv))

```


Nummber of pigs per Litter for 7 litters that has RNA samples
```{r }
ggplot2::ggplot(dat =num.sample, aes(x = dpv)) +
  ggplot2::geom_bar(aes(fill = Litters))+
  ggplot2::scale_x_continuous(breaks=unique(num.sample$dpv))

```

RINe for each pig over dpv

```{r }
ggplot2::ggplot(dat =num.sample, aes(x = dpv, y = RINe,group = PHGC_16_PigID, color = PHGC_16_PigID)) +
  ggplot2::geom_line()
  ggplot2::scale_x_continuous(breaks=unique(num.sample$dpv))

```

```{r}
num.sample[num.sample$RINe<4,c("RNAid", "RINe")]
sum(num.sample$RINe >=7)
```


## Explanatory Analysis of QuantSeq data
```{r}
library(DESeq2)

```

First filtering Quantseq data using threshold 0.5 and 3 nonzero counts.
```{r}
dat1 <- dat%>%filter(rowMeans(dat) >=.5 & rowSums(dat>0) >=3)

```

Normalize data using upper quartile. 
```{r}
offset <- apply(dat1, 2, quantile,  .75)

dat2 <- sweep(x = dat1, MARGIN = 2, STATS = exp(log(offset) - log(mean(offset))), 
              FUN = "/")


```

Multidensity function from DESeq2
```{r}
geneplotter::multidensity( dat2,
    xlab="mean counts", xlim=c(0, 1000))


```

multiecdf
```{r}

dat3 <- reshape2::melt(dat2)
head(dat3)
geneplotter::multiecdf(dat3$value~dat3$variable,xlab="mean counts", xlim=c(0, 1000))

```

PCA for top 500 genes showing the highest variance.
```{r}
library(limma)
ntop = nrow(dat2)
pcaplot <- function(timepoint = 4){
  
ncolum <- which(num.sample$dpv == timepoint)
  vout <- voom(counts = dat2[, ncolum], 
             lib.size = apply(dat2[, ncolum], 2, quantile, .75) )
Pvars <- genefilter::rowVars(vout$E)
select <- order(Pvars, decreasing = TRUE)[seq_len(min(ntop, 
        length(Pvars)))]
dataset <- vout$E[select,]
  pca <- prcomp(t(dataset), scale = F)
  percenvar <- round(100*pca$sdev^2/sum(pca$sdev^2),1)
  dataGG <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2],
                       percenvar = percenvar,
                       Name = colnames(dat2[, ncolum]),
                       WUR = num.sample$WUR[ncolum],
                       Vac = num.sample$VacStatue[ncolum],
                       Time = num.sample$dpv[ncolum], 
                       Litters = num.sample$Litters[ncolum])
  rownames(dataGG) <- num.sample$RNAid[ncolum]
  p <- ggplot(data = dataGG, aes(x = PC1, y = PC2, color = WUR, shape = Vac, label = Name)) + 
    geom_point(size = I(2))+ 
    ggplot2::ggtitle(paste0("Time point = ", timepoint, " dpv"))+
    ggplot2::xlab(label =paste0("PC1, VarExp:", round(percenvar[1],1) ))+
    ggplot2::ylab((label = paste0("PC2, VarExp:", round(percenvar[2],1))))
p
}

timevec <- sort(unique(num.sample$dpv))
timevec
ggplotly(pcaplot(timepoint = timevec[1]))
```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[1]], 2, sum),las = 2, main = paste0("dpv = ", timevec[1]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[2]))

```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[2]], 2, sum),las = 2, main = paste0("dpv = ", timevec[2]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[3]))


```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[3]], 2, sum),las = 2, main = paste0("dpv = ", timevec[3]))

```

```{r}

ggplotly(pcaplot(timepoint = timevec[4]))



```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[4]], 2, sum),las = 2, main = paste0("dpv = ", timevec[4]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[5]))


```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[5]], 2, sum),las = 2, main = paste0("dpv = ", timevec[5]))

```

```{r}

ggplotly(pcaplot(timepoint = timevec[6]))


```

Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[6]], 2, sum),las = 2, main = paste0("dpv = ", timevec[6]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[7]))


```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[7]], 2, sum),las = 2, main = paste0("dpv = ", timevec[7]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[8]))


```

Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[8]], 2, sum),las = 2, main = paste0("dpv = ", timevec[8]))

```
```{r}

ggplotly(pcaplot(timepoint = timevec[9]))


```
Check outliers samples.
```{r}
barplot(apply(dat1[, num.sample$dpv==timevec[9]], 2, sum),las = 2, main = paste0("dpv = ", timevec[9]))

```


```{r}

ggplotly(pcaplot(timepoint = timevec[10]))


```

Check outlier samples.
```{r}

barplot(apply(dat1[, num.sample$dpv==timevec[10]], 2, sum),las = 2, main = paste0("dpv = ", timevec[10]))
```


Save process data to data folder.
```{r}
library(plyr)
out <- plyr::llply(timevec, function(timepoint){
  counts <- dat[, num.sample$dpv==timepoint]
  counts <- counts[rowMeans(counts) >2 & rowSums(counts > 0) >2, ]
  covset <- num.sample %>% 
    dplyr::filter(dpv == timepoint) %>% 
    dplyr::select(RNAid, Litters, WUR, VacStatue, 
                  PHGC_16_PigID, RINe, Plate, WUR_Vac) %>% 
    dplyr::mutate(RNAid = as.character(RNAid), 
                  Litters = as.factor(as.character(Litters)),
                  WUR = as.factor(as.character(WUR)),
                  VacStatue = as.factor(as.character(VacStatue)),
                  PHGC_16_PigID = as.factor(as.character(PHGC_16_PigID)),
                  Plate = as.factor(as.character(Plate)),
                  WUR_Vac = as.factor(as.character(WUR_Vac))
                  )
  datout <- list(counts = counts, covset = covset)
  saveRDS(datout, file = paste0("../data/dat_", timepoint, ".rds"))
  datout
  
})
names(out) <- paste0("dat_", timevec)
saveRDS(out, file = paste0("../data/datall.rds"))


```


